Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Edits to Section 6.

Research software is critical for tackling problems in areas as diverse as
manufacturing, financial planning, environmental policy and medical diagnosis
and treatment.  However, developing reliable, reproducible, sustainable and fast
research software to address new problems is challenging because of the
complexity of the physical models and the nuances of floating point and parallel
computation. The importance of research software and the difficulty with its
development have prompted multiple researchers to investigate how developing
this software differs from other classes of software.  Previous studies have
focused on surveying developers
\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
\cite{SandersAndKelly2008} and case studies \cite{CarverEtAl2007,Segal2005}.  A
valuable source of information that has received less attention is the data in
publicly available software repositories.  We propose digging deeper into how
well projects are applying best practices by using a new methodology that
combines manual and automated techniques to extract repository-based
information.

@@

Continuous integration is rarely used in LBM (3 of 24 packages or 12.5\%). This
contrasts with the frequency with which continuous integration is recommended in
research software development guidelines
\cite{BrettEtAl2021,Brown2015,ThielEtAl2020,Zadka2018,vanGompelEtAl2016}. In
this case, it seems likely that the recommendations are ahead of common
practice.

@@

This section answers research question \rqref{RQ_CompareMethodologies} by
comparing the principles, processes and methodologies used for LBM software to
what can be gleaned from the literature on research software in general. In our
data collection for LBM software, the software development process is not
explicitly indicated in the artifacts for most of the packages. However, during
our interviews one developer (ESPResSo) told us their non-rigorous development
model is like a combination of agile and waterfall. Employing a loosely defined
process makes sense for LBM software, given that the teams are generally small
and self-contained.  Although eleven of the packages explicitly convey that they
would accept outside contributors, generally the teams are centralized, often
working at the same institution.  Working at the same institution means that an
informal process can show success, since informal conversations are relatively
easy to have.

Interviews with developers confirmed a similar project management processes. In
teams of only a couple of developers, additions of new features or major changes
are discussed with the entire team. Projects with more than a couple developers
have lead developer roles. These lead developers review potential additions to
the software. One of the developers (ESPResSo) that was interviewed noted that
an ad hoc peer review process is used to assess major changes and additions.
Using peer review (also called technical review) matches with recommended
practice for research software \cite{HerouxEtAl2008,Givler2020,OrvizEtAl2017,USGS2019}.

Two types of software changes were discussed during interviews with developers.
One is feature additions, which arise from a scientific or functional need.
These changes involve formal discussions within the development team, and lead
developer participation is mandatory. The other change type is code refactoring,
which only sometimes involves formal discussions with the development team. New
developers were noted to play an increased role in these changes compared to the
former changes. Software bugs are typically addressed in a similar fashion as
code refactoring.  Issue tracking is commonly used to manage these changes.

Our observations of a informally defined process, with elements of agile
methods, matches what has been observed for research software in general.
Scientific developers naturally use an agile philosophy \cite{AckroydEtAl2008,CarverEtAl2007,EasterbrookAndJohns2009,Segal2005,HeatonAndCarver2015}, or an
amethododical process \cite{Kelly2013}, or a knowledge acquisition driven
process \cite{Kelly2015}.

Most of the software packages do not explicitly state the motivations or design
principles that were considered when developing the software. One package,
Sailfish, indicates in its artifacts that shortening the development time was
considered in early stages of design, with the developers using Python and
CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
are explicitly listed as performance, scalability, agility and extendability,
maintenance, and ease of use. The project scored well in these categories during
our assessment.  The quality priorities for Sailfish roughly match the
priorities observed for research software in general.
\cite{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
qualities, in decreasing order of importance: reliability, functionality,
maintainability, availability, performance, flexibility, testability, usability,
reusability, traceability, and portability. The Sailfish list does not list
reliability or functionality, but we can safely assume those are implicitly high
priorities for any scientific project.  In earlier studies
\cite{KellyAndSanders2008} and \cite{CarverEtAl2007} both highlight how
important correctness is for research software.

During our interviews, documentation was noted as playing a significant role in
the development process, specifically with on-boarding new developers. A goal of
documentation is to lower the entry barrier for these new contributors. The
documentation provides information on how to get started, orients the user to
artifacts and the source code, and explains how the system works, including the
so-called simulation engine and interface.  This emphasis on documentation,
especially for new developers, is echoed in research software guidelines.
Multiple guidelines recommend a document explaining how to contribute to a
project, often named CONTRIBUTING \cite{Yo2021,BrettEtAl2021,WilsonEtAl2016,ThielEtAl2020,vanGompelEtAl2016,OrvizEtAl2017,FLOSS2022,JimenezEtAl2017}.
Tutorials \cite{ThielEtAl2020}, trouble shooting guides \cite{OrvizEtAl2017,SSI2022} and quick start examples \cite{ThielEtAl2020,vanGompelEtAl2016} are
also recommended.  \cite{SmithAndRoscoe2018} suggests including instructions
specifically for on-boarding new developers. For open source software in general
(not just research software), \cite{Fogel2005} recommends providing tutorial
style examples, developer guidelines, demos and screenshots.

>>>>

Research software is critical for tackling problems in areas as diverse as
manufacturing, financial planning, environmental policy and medical diagnosis
and treatment.  However, developing reliable, reproducible, sustainable and fast
research software to address new problems is challenging because of the
complexity of the physical models and the nuances of floating point and parallel
computation. The importance of research software and the difficulty with its
development have prompted multiple researchers to investigate how developing
this software differs from other classes of software.  Previous studies have
focused on surveying developers
\cite{HannayEtAl2009,Nguyen-HoanEtAl2010,PintoEtAl2018}, developer interviews
\cite{Kelly2013} and case studies \cite{CarverEtAl2007,Segal2005}.  A
valuable source of information that has received less attention is the data in
publicly available software repositories.  We propose digging deeper into how
well projects are applying best practices by using a new methodology that
combines manual and automated techniques to extract repository-based
information.

@@

Continuous integration is rarely used in LBM (3 of 24 packages or 12.5\%). This
contrasts with the frequency with which continuous integration is recommended in
research software development
guidelines~\cite{BrettEtAl2021,vanGompelEtAl2016,ThielEtAl2020}. In this case,
it seems likely that the recommendations are ahead of common practice.

@@

The interview data on development processes is used to answer research question
\rqref{RQ_CompareMethodologies}.  This data should be contrasted with the
development process used by research software in general. The literature
suggests that scientific developers naturally use an agile philosophy
\cite{CarverEtAl2007,Segal2005}, or an amethododical process \cite{Kelly2013}.
Another point of comparison should be on the use of peer review, since peer
review is frequently recommended in software development
guidelines~\cite{HerouxEtAl2008,OrvizEtAl2017,USGS2019}.

The running example of LBM confirms an informal process, with elements of agile
methods. The software development process is not explicitly indicated in the
artifacts for most of the packages. However, during interviews one developer
(ESPResSo) told us their non-rigorous development model is like a combination of
agile and waterfall. Employing a loosely defined process makes sense for LBM
software, given that the teams are generally small and self-contained. One of
the developers (ESPResSo) that was interviewed also noted that an ad hoc peer
review process is used to assess major changes and additions, matching the
common recommendation.
